---
title: "Analysis with Mpath"
output: html_notebook
---

```{r}
library(Mpath)
```

Quality check
-------------

```{r}
# QC_gene removes genes that have TPM values < 1
# in more than 95 percent of cells in each group

rpkmFile   <- "mpath/only_genes_CD4tcells_tpm.txt"
rpkmQCFile <- "mpath/only_genes_CD4tcells_tpm_geneQC.txt"
sampleFile <- NULL
QC_gene(rpkmFile   = rpkmFile,
        rpkmQCFile = rpkmQCFile,
        sampleFile = sampleFile, threshold = 0.05, method = "any")
```


```{r}
rpkmQC <- fread(rpkmQCFile)
dim(rpkmQC)
```


```{r}

write.table(sampleInfo, file="mpath/CD4tcells_sample_day_info.txt", sep="\t", 
            row.names=F, col.names=T, quote=F)
```


```{r}
rpkmFile   <- "mpath/only_genes_CD4tcells_tpm_geneQC.txt"
sampleFile <- "mpath/CD4tcells_sample_day_info.txt"
find_optimal_cluster_number(rpkmFile = rpkmFile,
                            sampleFile = sampleFile,
                            min_cluster_num = 7, max_cluster_num = 15,
                            diversity_cut = 0.6, size_cut = 0.05)
### optimal number of cluster -> 11
```

### Landmark designation

```{r}
rpkmFile   <- "mpath/only_genes_CD4tcells_tpm_geneQC.txt"
baseName   <- "mpath/only_genes_CD4tcells_tpm_geneQC";
sampleFile <- "mpath/CD4tcells_sample_day_info.txt"
landmark_cluster <- landmark_designation(rpkmFile = rpkmFile,
                                         baseName = baseName,
                                         sampleFile = sampleFile,
                                         method = "diversity_size",
                                         numcluster = 11, diversity_cut=0.6,
                                         size_cut=0.05)
```

```{r}
landmark_cluster2 <- landmark_cluster
landmark_cluster2$cell <- gsub("^X", "",   landmark_cluster2$cell)
landmark_cluster2$cell <- gsub("\\.", "-", landmark_cluster2$cell)

write.table(landmark_cluster2, file="mpath/only_genes_CD4tcells_tpm_geneQC_landmark_cluster2.txt",
            row.names=F, col.names = T, sep="\t")
```


### Plot hierachical clustering 

```{r}
SC_hc_colorCode(dataFile   = "mpath/only_genes_CD4tcells_tpm_geneQC.txt",
                cuttree_k  = 11,
                sampleFile = "mpath/CD4tcells_sample_day_info.txt",
                width = 22, height = 10, iflog2 = TRUE,
                colorPalette = NULL)
```

### Construct weighted neighborhood network

```{r}

exprs    <- "mpath/only_genes_CD4tcells_tpm_geneQC.txt"
baseName <- "mpath/only_genes_CD4tcells_tpm_geneQC"
neighbor_network <- build_network(exprs = exprs,
                                  landmark_cluster = landmark_cluster2,
                                  baseName = baseName)
```

### TrimNet: trim edges of lower weights

```{r}
baseName = "mpath/only_genes_CD4tcells_tpm_geneQC";
trimmed_net <- trim_net(neighbor_network, textSize=30,
                        baseName = baseName,
                        method = "mst")
```

### plot trimmed net and color-code the nodes by gene expression

```{r}
rpkmFile <- "mpath/only_genes_CD4tcells_tpm_geneQC.txt"
lmFile   <- "mpath/only_genes_CD4tcells_tpm_geneQC_landmark_cluster.txt"
color_code_node_2(networkFile = trimmed_net,
                  rpkmFile = rpkmFile,
                  lmFile = lmFile,
                  geneName=c("Irf8","Id2","Batf3"),
                  baseName="cDC1_marker",
                  seed=NULL)
```

### Re-order the cells on the path connecting landmark 
### "day2_4", "day2_7",  "day4_2", "day4_9", "day4_5",
### "day4_8", "day7_10", "day7_11"

```{r}
exprs  <- "mpath/only_genes_CD4tcells_tpm_geneQC.txt"
ccFile <- "mpath/only_genes_CD4tcells_tpm_geneQC_landmark_cluster2.txt"
order2 <- nbor_order(exprs = exprs,
                    ccFile = ccFile, 
                    lm_order = c("day_2_4", "day_2_7",  "day_4_2", "day_4_9", "day_4_5",
                                 "day_4_8", "day_7_10", "day_7_11"),
                    if_bb_only=FALSE,
                    method=1)
```

### identify genes that changed along the cell re-ordering

```{r}
deg <- vgam_deg(exprs = "mpath/only_genes_CD4tcells_tpm_geneQC.txt",
                order = order2,
                lm_order = c("day_2_4", "day_2_7",  "day_4_2", "day_4_9", "day_4_5",
                                 "day_4_8", "day_7_10", "day_7_11"),
                min_expr=1,
                p_threshold=0.05)
```

### plot heatmap of genes that changed along the cell re-ordering

```{r}
dev.off()
heatmap_nbor(exprs = "mpath/only_genes_CD4tcells_tpm_geneQC.txt",                        
             cell_order = "mpath/day_2_4_day_2_7_day_4_2_day_4_9_day_4_5_day_4_8_day_7_10_day_7_11_order.txt", 
             plot_genes = "mpath/day_2_4_day_2_7_day_4_2_day_4_9_day_4_5_day_4_8_day_7_10_day_7_11_vgam_deg0.05.txt",
             cell_annotation = "mpath/CD4tcells_sample_day_info.txt", 
             num_gene_cluster = 6,
             hm_height = 15, hm_width = 10,
             baseName = "mpath/day_2_4_day_2_7_day_4_2_day_4_9_day_4_5_day_4_8_day_7_10_day_7_11")
```

```{r}
mycolors <- c("#81CDC3", "#EB867F",  "#D0AB80", "#BDD68F", "#F9EF76")
names(mycolors) <- c("day_0", "day_2", "day_3", "day_4", "day_7")
```


```{r}
### path 1: "day_2_4", "day_2_7",  "day_4_2", "day_4_9", "day_4_5", "day_4_8"
ccFile <- "mpath/only_genes_CD4tcells_tpm_geneQC_landmark_cluster2.txt"
order1  <- nbor_order(exprs = "mpath/only_genes_CD4tcells_tpm_geneQC.txt",
                     ccFile = ccFile, 
                     lm_order = c("day_2_4", "day_2_7",  "day_4_2", "day_4_9", "day_4_5",
                                 "day_4_8"),
                     if_bb_only = TRUE,
                     method = 1)

deg <- vgam_deg(exprs = "mpath/only_genes_CD4tcells_tpm_geneQC.txt",
                order = order1,
                lm_order =  c("day_2_4", "day_2_7",  "day_4_2", "day_4_9", "day_4_5", "day_4_8"),
                min_expr = 1,
                p_threshold = 0.05)

heatmap_nbor(exprs = "mpath/only_genes_CD4tcells_tpm_geneQC.txt",                        
            cell_order = "day_2_4_day_2_7_day_4_2_day_4_9_day_4_5_day_4_8_order.txt",
            plot_genes = "day_2_4_day_2_7_day_4_2_day_4_9_day_4_5_day_4_8_vgam_deg0.05.txt",
            cell_annotation = "mpath/CD4tcells_sample_day_info2.txt", 
            num_gene_cluster = 6,
            hm_height = 15, hm_width = 10,
            baseName = "mpath/CD4tcells_path1_order_vgam_deg0.05",
            colorPalette = mycolors)
```



```{r}
### path 2: "day_2_4", "day_2_7",  "day_4_2", "day_7_10", "day_7_11"
ccFile <- "mpath/only_genes_CD4tcells_tpm_geneQC_landmark_cluster2.txt"
order2 <- nbor_order(exprs = "mpath/only_genes_CD4tcells_tpm_geneQC.txt",
                    ccFile = ccFile, 
                    lm_order = c("day_2_4", "day_2_7",  "day_4_2", "day_7_10", "day_7_11"),
                    if_bb_only = TRUE,
                    method=1)

deg <- vgam_deg(exprs = "mpath/only_genes_CD4tcells_tpm_geneQC.txt",
               order = order2,
               lm_order = c("day_2_4", "day_2_7",  "day_4_2", "day_7_10", "day_7_11"),
               min_expr=1,
               p_threshold=0.05)

heatmap_nbor(exprs = "mpath/only_genes_CD4tcells_tpm_geneQC.txt",                        
            cell_order = "day_2_4_day_2_7_day_4_2_day_7_10_day_7_11_order.txt",
            plot_genes = "day_2_4_day_2_7_day_4_2_day_7_10_day_7_11_vgam_deg0.05.txt",
            cell_annotation = "mpath/CD4tcells_sample_day_info2.txt", 
            num_gene_cluster = 6,
            hm_height = 15, hm_width = 10,
            baseName = "mpath/CD4tcells_path2_order_vgam_deg0.05",
            colorPalette = mycolors)
```


```{r}
deg1 <- read.table("day_2_4_day_2_7_day_4_2_day_4_9_day_4_5_day_4_8_vgam_deg0.05.txt",sep="\t",header=T)
deg2 <- read.table("day_2_4_day_2_7_day_4_2_day_7_10_day_7_11_vgam_deg0.05.txt",sep="\t",header=T)
deg <- unique(c(as.character(deg1[,1]),as.character(deg2[,1])))

heatmap_nbor(exprs = "mpath/only_genes_CD4tcells_tpm_geneQC.txt",                   
            cell_order = c(order1,order2), 
            plot_genes = deg,
            cell_annotation = "mpath/CD4tcells_sample_day_info2.txt", 
            num_gene_cluster = 7,
            hm_height = 15, hm_width = 10,
            baseName = "mpath/Path12_method1orderedbackbone_progression_heatmap", 
            n_linechart = list(order1,order2),
          colorPalette = mycolors)
             
```

```{r}
geneMap <- fread("/nfs/research2/teichmann/reference/mus-musculus/ensembl85-mouse-tx-annotation.csv")
geneMap <- subset(geneMap, select=c("Ensembl Gene ID", "Associated Gene Name"))
geneMap <- unique(geneMap)
```


```{r}
deg1 <- merge(deg1, geneMap, by.x="X", by.y="Ensembl Gene ID", all.x=T)
deg1 <- deg1[order(deg1$p.adj),]
head(deg1, n=20)
```

```{r}
deg2 <- merge(deg2, geneMap, by.x="X", by.y="Ensembl Gene ID", all.x=T)
deg2 <- deg2[order(deg2$p.adj),]
head(deg2, n=20)
```

